---
title: Problem set 4
author: Camila R. Guetter
date: 2025-10-05
---

In the next problem set, we plan to explore the relationship between COVID-19 death rates and vaccination rates across US states by visually examining their correlation. This analysis will involve gathering COVID-19 related data from the CDC's API and then extensively processing it to merge the various datasets. Since the population sizes of states vary significantly, we will focus on comparing rates rather than absolute numbers. To facilitate this, we will also source population data from the US Census to accurately calculate these rates. 

In this problem set we will learn how to extract and wrangle data from the data US Census and CDC APIs.

1. Get an API key from the US Census at <https://api.census.gov/data/key_signup.html>. You can't share this public key. But your code has to run on a TFs computer. Assume the TF will have a file in their working directory named `census-key.R` with the following one line of code:

```
census_key <- "A_CENSUS_KEY_THAT_WORKS"
```

Write a first line of code for your problem set that defines `census_key` by running the code in the file `census-key.R`.

```{r}
# Source census key 
source("census-key.R")

```

2. The [US Census API User Guide](https://www.census.gov/content/dam/Census/data/developers/api-user-guide/api-user-guide.pdf)
provides details on how to leverage this valuable resource. We are interested in vintage population estimates for years 2021 and 2022. From the documentation we find that the _endpoint_ is:


```{r}
url <- "https://api.census.gov/data/2021/pep/population"
```

Use the **httr2** package to construct the following GET request.

```
https://api.census.gov/data/2021/pep/population?get=POP_2020,POP_2021,NAME&for=state:*&key=YOURKEYHERE
```

Create an object called `request` of class `httr2_request` with this URL as an endpoint.
Hint: Print out `request` to check that the URL matches what we want.

```{r}
#| message: false
#| warning: false
library(httr2)

#GET request (for 2020-2021)
request <- request("https://api.census.gov/data/2021/pep/population") |>
  req_url_query(
    get = "POP_2020,POP_2021,NAME",
    `for` = "state:*",
    key = census_key
  )

# Print out 'request' (here commented out, to avoid printing my API key)
#request
```

3. Make a request to the US Census API using the `request` object. Save the response to and object named `response`. Check the response status of your request and make sure it was successful. You can learn about _status codes_ [here](https://en.wikipedia.org/wiki/List_of_HTTP_status_codes).

```{r}
# Perform the request and save to 'response' object
response <- request |> 
  req_perform()

# Check status of request
resp_status(response) #200 means ok
```

4. Use a function from the **httr2** package to determine the content type of your response.

```{r}
# Check the content type of the response
resp_content_type(response)
```

5. Use just one line of code and one function to extract the data into a matrix. 
Hints: 1) Use the `resp_body_json` function. 2) The first row of the matrix will be the variable names and this OK as we will fix in the next exercise.

```{r}
# Extract data into matrix
population <- resp_body_json(response, simplifyVector = TRUE)

```

6. Examine the `population` matrix you just created. Notice that 1) it is not tidy, 2) the column types are not what we want, and 3) the first row is a header. Convert `population` to a tidy dataset. Remove the state ID column and change the name of the column with state names to `state_name`. Add a column with state abbreviations called `state`. Make sure you assign the abbreviations for DC and PR correctly. Hint: Use the **janitor** package to make the first row the header. 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)

population <- population |>
  as_tibble(.name_repair = "minimal") |>         # convert matrix to tibble
  row_to_names(row_number = 1) |>                # change first row to header
  rename(state_name = any_of(c("NAME", "name"))) |>   # rename state name col
  select(-any_of(c("state", "STATE"))) |>        # remove state column
  pivot_longer(                                  
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = as.integer(sub("^POP_", "", year)),   # remove "POP_" from year
    population = as.numeric(population)          # parse columns to numeric
  ) |>
  left_join(                                     # add state abbreviations
    tibble(state_name = state.name, state = state.abb),
    by = "state_name"
  ) |>
  mutate(                                         # abbreviations for DC & PR
    state = case_when(
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico"          ~ "PR",
      TRUE                                 ~ state
    )
  ) |>
  relocate(state, .after = state_name)
```

7. As a check, make a barplot of states' 2021 and 2022 populations. Show the state names in the y-axis ordered by population size.  Hint: You will need to use `reorder` and use `facet_wrap`.

```{r, fig.height=12, fig.width=8}

population |>
  filter(year %in% c(2020, 2021)) |>
  ggplot(aes(
    x = reorder(state_name, population),
    y = population,
    fill = factor(year)
  )) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ year, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  labs(
    x = "State",
    y = "Population",
    fill = "Year",
    title = "State Populations in 2020 and 2021"
  ) +
  theme_minimal()
```


8. The following URL:

```{r}
url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
```

points to a JSON file that lists the states in the 10 Public Health Service (PHS) defined by CDC. We want to add these regions to the `population` dataset. To facilitate this create a data frame called `regions` that has two columns `state_name`, `region`, `region_name`. One of the regions has a long name. Change it to something shorter.

```{r}
#| message: false
#| warning: false
library(jsonlite)
library(purrr)

# Parse JSON from the URL
regions_raw <- fromJSON(url, simplifyVector = FALSE)

# Work on data frame
regions <- regions_raw |>
  map_df(~ tibble(
    region = as.integer(.x[[1]]),
    region_name = .x[[2]],
    states = list(.x[[3]])
  )) |>
  unnest(cols = states, names_repair = "minimal") |>  
  rename(state_name = states) |>
  mutate(
    region_name = ifelse(
      region_name == "New York and New Jersey, Puerto Rico, Virgin Islands",
      "NY/NJ/PR/VI",
      region_name
    )
  )

# Go from 59 to 52 rows (select states)
valid_states <- c(state.name, "District of Columbia", "Puerto Rico")
regions <- regions[regions$state_name %in% valid_states, , drop = FALSE]

# Reorder columns to follow question prompt
regions <- regions[, c("state_name", "region", "region_name")]

# Check result
#nrow(regions)    # now is 52 in total
#head(regions)

```

9. Add a region and region name columns to the `population` data frame.

```{r}
# Change from list-column to a character vector (so that I can merge later)
regions$state_name <- as.character(unlist(regions$state_name, use.names = FALSE))

# Check
#str(regions$state_name) 

population <- population |>
  left_join(regions, by = "state_name")
```


10. From reading <https://data.cdc.gov/> we learn the endpoint `https://data.cdc.gov/resource/pwn4-m3yp.json` provides state level data from SARS-COV2 cases. Use the **httr2** tools you have learned to download this into a data frame. Is all the data there? If not, comment on why.


```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

# Create and perform the request
cases_raw <- request(api) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

# Check data
str(cases_raw)
nrow(cases_raw)
```

*Answer: No, not all the data is there. The CDC, has a default limit to 1000 rows as explained in the next exercise, so you need to set a new limit if you want to override that defauly limit of rows when getting the data.*

We see exactly 1,000 rows. We should be seeing over $52 \times 3$ rows per state.

11. The reason you see exactly 1,000 rows is because CDC has a default limit. You can change this limit by adding `$limit=10000000000` to the request. Rewrite the previous request to ensure that you receive all the data. Then wrangle the resulting data frame to produce a data frame with columns `state`, `date` (should be the end date) and `cases`. Make sure the cases are numeric and the dates are in `Date` ISO-8601 format.

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json?$limit=10000000000"

cases_raw <- request(api) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

cases <- data.frame(
  state = cases_raw$state,
  date  = as.Date(cases_raw$end_date),
  cases = as.numeric(cases_raw$new_cases),
  stringsAsFactors = FALSE
)
```


12. For 2020 and 2021, make a time series plot of cases per 100,000 versus time for each state. Stratify the plot by region name. Make sure to label you graph appropriately. 

```{r}

# Filter for years 2020–2021, merge datasets, calculate cases per 100,000
cases_pop <- cases |>
  filter(format(date, "%Y") %in% c("2020", "2021")) |>
  mutate(year = as.integer(format(date, "%Y"))) |>
  left_join(population |> select(state, state_name, year, population),
            by = c("state", "year")) |>
  left_join(regions |> select(state_name, region_name),
            by = "state_name") |>
  mutate(cases_per_100k = (cases / population) * 1e5) |>
  filter(!is.na(region_name))   # removes rows with missing region info (to avoid NA plot)

# Plot
ggplot(cases_pop, aes(x = date, y = cases_per_100k, group = state, color = state)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  facet_wrap(~ region_name, scales = "free_y") +
  labs(
    title = "Daily SARS-COV2 Cases per 100,000 Population \n Plotted by State and Stratified by US Region (2020–2021)",
    x = "Date",
    y = "New Cases per 100,000",
    color = "State"
  ) +
  guides(color = "none") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 13),  
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),       
    strip.text = element_text(face = "bold"),                          
    panel.spacing = unit(0.8, "lines"),                                
    plot.title.position = "panel"                                      
  )

```

13. The dates in the `cases` dataset are stored as character strings. Use the **lubridate** package to properly parse the `date` column, then create a summary table showing the total COVID-19 cases by month and year for 2020 and 2021. The table should have columns for year, month (as month name), and total cases across all states. Order by year and month. Use the **knitr** package and `kable()` function to display the results as a formatted table.

```{r}
library(lubridate)
library(tidyverse)
library(knitr)

cases |>
  mutate(date = ymd(date)) |>
  filter(year(date) %in% c(2020, 2021)) |>
  mutate(
    year  = year(date),
    month = month(date, label = TRUE, abbr = FALSE)
  ) |>
  group_by(year, month) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(year, month) |>
  mutate(year = as.character(year)) |>   # to avoid commas in year rows
  rename(
    "Year" = year,
    "Month" = month,
    "Total Cases" = total_cases
  ) |> 
  kable(
    caption = "**Total COVID-19 Daily New Cases by Month Across All States (2020–2021)**",
    digits = 0,
    format.args = list(big.mark = ",")
  )

```

14. The following URL provides additional COVID-19 data from the CDC in JSON format:

```{r}
deaths_url <- "https://data.cdc.gov/resource/9bhg-hcku.json?$limit=10000000000"
```

Use **httr2** to download COVID-19 death data from this endpoint. Make sure to remove the default limit to get all available data. Create a clean dataset called `deaths` with columns `state`, `date`, and `deaths` (renamed from the original column name). Ensure dates are in proper Date format and deaths are numeric.


```{r}
# Request and download JSON data
deaths_raw <- request(deaths_url) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

# Define state names to match previous datasets (50 states + DC + PR)
valid_full <- c(state.name, "District of Columbia", "Puerto Rico")

# Create clean dataset
deaths <- deaths_raw |>
  as_tibble() |>
  select(
    state,
    date   = end_date,             
    deaths = covid_19_deaths       
  ) |>
  mutate(
    date   = as.Date(date),       
    deaths = as.numeric(deaths)    
  ) |>
  filter(
    state %in% valid_full,        
    !is.na(date),
    !is.na(deaths)
  )

# Check
#glimpse(deaths)

```

15. Using the `deaths` dataset you created, make a bar plot showing the total COVID-19 deaths by state. Show only the top 10 states with the highest death counts. Order the bars from highest to lowest and use appropriate labels and title.

```{r}
# Summarize total deaths by state
top10_deaths <- deaths |>
  group_by(state) |>
  summarise(total_deaths = sum(deaths, na.rm = TRUE)) |>
  arrange(desc(total_deaths)) |>
  slice_head(n = 10)  # keep top 10 states for plotting

# Plot
ggplot(top10_deaths, aes(x = reorder(state, total_deaths), y = total_deaths)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 U.S. States by Total COVID-19 Deaths",
    x = "State",
    y = "Total Deaths"
  ) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 10)
  )
```
